The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2016-08-23 script/en/chapters/cryptomethods.tex
CT Book ed 12: Major update -- many changes, corrections, and additions during last 4 months

% $Id$
% ..............................................................................
%                V E R S C H L U E S S E L U N G S V E R F A H R E N
%
% Schreibregelung / Syntax rules English:
%         - Chapter header with all nouns/verbs capitalized.
%         - All other headers in the normal lower/upper case manner
%           like sentences, but without dot at the end.
%
% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

\hypertarget{Kapitel_1}{}
\chapter{Security Definitions and Encryption Procedures}
\label{Label_Kapitel_1}
(Bernhard Esslinger, Joerg-Cornelius Schneider, May 1999; Updates Dec. 2001, Feb. 2003, June 2005, July 2007, Jan. 2010, Mar. 2013, Mar. 2016)

@@

The purpose of encryption \index{encryption} is to change data in such a way
that only an authorized recipient is able to reconstruct the plaintext. This
allows us to transmit data without worrying about it getting into unauthorized
hands. Authorized recipients possess a piece of secret information -- called the
key -- which allows them to decrypt the data while it remains hidden from
everyone else.%\par \vskip + 3pt

For explanations in the following we use the notation from 
Figure \ref{Generic-Notations-when-Encrypting}:
\begin{figure}[ht]
\begin{center}
\includegraphics[scale=0.7]{figures/Generic-Notation-Encryption_en.png}
\caption{Common notations when using ciphers} 
\label{Generic-Notations-when-Encrypting}
\end{center}
\end{figure}

@@

% --------------------------------------------------------------------------
\hypertarget{cm_Section_Security_Definitions}{}
\section[Security definitions]{Security definitions}
\label{cm_Section_Security_Definitions}
\index{security definitions}

@@

\vskip +3pt
Depending on the adversary's capabilities there are mainly two basic notations
of security distinguished in literature
(see e.g. {\em Contemporary Cryptography} \cite{cm:Oppliger2011}):

@@

\item {\bf Computational, conditional or practical security}\\
A cipher is computationally secure if it is theoretically possible to break such
a system but it is infeasible to do so by any known practical means. Theoretical
advances (e.g., improvements in integer factorization algorithms) and faster
computing technology require these solutions to be continually adapted.

@@

\item {\bf Information-theoretical or unconditional security}\\
A cipher is unconditionally secure if it is secure no matter how much resources
(time, space) the attacker has. So even in the case where an adversary has
unlimited resources for solving a problem e.g. breaking a cipher. An adversary
has not sufficient information for doing that because he is unable to gain any
meaningful data from a ciphertext.

There exist information-theoretically secure schemes that provably cannot be
broken even with unlimited computing power -- an example is the
{\em one-time pad}\index{one-time pad} (OTP\index{OTP}).

@@

\item {\bf Ad-hoc security}
A cryptographic system has this security feature if it is not worth to try to
break into such a system because of inadequate price of data with comparison to
price of work needed to do so. Or an attack can't be done in sufficiently short
time (see {\em Handbook of Applied Cryptography} \cite{cm:Menezes2001}).

@@

\vskip +3pt
For good procedures used today the time taken to break them is so long that
it is practically impossible to do so, and these procedures can therefore be
considered (practically) secure -- from a pure algorithm's point of view.%\par

\vskip +3pt
We basically distinguish between symmetric and asymmetric encryption
procedures.
The book of Bruce Schneier \cite{cm:Schneier1996cm} offers a good overview of
the different encryption algorithms.
\vskip +3pt

@@

@@

% --------------------------------------------------------------------------
\section[Symmetric encryption]
{Symmetric encryption\footnotemark}
  \footnotetext{%
    With CrypTool 1 ({\bf CT1})\index{CT1}\index{CrypTool 1} you can execute
    the following modern symmetric encryption algorithms 
    (using the menu path {\bf Crypt \textbackslash{} Symmetric (modern)}):\\
    IDEA, RC2, RC4, DES (ECB), DES (CBC), Triple-DES (ECB), Triple-DES (CBC),
    MARS (AES candidate), RC6 (AES candidate), Serpent (AES candidate), 
    Twofish (AES candidate), Rijndael (official AES algorithm)\index{AES}.\\
    With CrypTool 2 ({\bf CT2})\index{CT2}\index{CrypTool 2} you can execute the
    following modern symmetric encryption algorithms (using in the Startcenter
    {\bf Templates
    \textbackslash{} Cryptography \textbackslash{} Modern \textbackslash{}
    Symmetric}):\\
    AES, DES, PRESENT, RC2, RC4, SDES, TEA, Triple-DES, Twofish.\\
    In JCrypTool ({\bf JCT})\index{JCT}\index{JCrypTool} you can execute the
    following modern symmetric encryption algorithms:\\ 
    AES, Rijndael, Camellia, DES, Dragon, IDEA, LFSR, MARS, Misty1, RC2, RC5,
    RC6, SAFER+, SAFER++, Serpent, Shacal, Shacal2, Twofish.
  }

@@

This is visualized in Figure \ref{Figure_Symmetric-Enc_Secret-Key-Enc}:
\begin{figure}[ht]
\begin{center}
\includegraphics[scale=0.7]{figures/SymmetricEnc_Figure_Chap1_en.png}
\caption{Symmetric or Secret-Key Encryption} 
\label{Figure_Symmetric-Enc_Secret-Key-Enc}
\end{center}
\end{figure}

All classical ciphers are of the type symmetric. Examples can be found within
the CT programs, in chapter~\ref{Kapitel_PaperandPencil}
(``\nameref{Kapitel_PaperandPencil}'')
of this script, or in \cite{cm:Nichols1996}.
In this section however, we want to consider only modern symmetric mechanisms.

@@

Before AES, the most well-known symmetric encryption procedure was the \index{DES} DES algorithm.
The DES algorithm has been developed by IBM in collaboration with the
National Security Agency \index{NSA} (NSA), and was published as a standard in
1975. Despite the fact that the procedure is relatively old, no effective attack
on it has yet been detected. The most effective way of attacking consists of
testing (almost) all possible keys until the right one is found ({\em brute-force-attack}).
\index{attack!brute-force} Due to the relatively short key length of
effectively 56 bits (64 bits, which however include 8 parity bits), numerous
messages encrypted using DES have in the past been broken. Therefore, the
procedure can not be considered secure any longer. Alternatives to the DES
procedure include IDEA\index{IDEA}, Triple-DES and especially AES.
\par \vskip + 3pt

@@

More details and further references about the AES algorithms and the AES
candidates of the last round can be found i.e. within the online help of
CrypTool\index{CrypTool}.%
\footnote{%
      CrypTool 1 online help\index{CrypTool 1}: The index head-word {\bf AES}
      leads to the 3 help pages: {\bf AES candidates}, 
      {\bf The AES winner Rijndael} and 
      {\bf The Rijndael encryption algorithm}\\
      A comprehensive description of AES including C code can be found
      in \cite{cm:Haan2008}.
  }

% HACK to fix warning "destination with the same identifier .. has already been used, ...":
%\makeatletter \renewcommand{\thepage}{\csname @arabic\endcsname \c@page} \makeatother
% --------------------------------------------------------------------------
\subsection{Results about cryptanalysis of AES}
\label{NeueAES-Analyse}
\index{cryptanalysis}
\index{AES}

Below you will find some results, which have recently called into question the security of the AES algorithm -- from our point of view these doubts practically still remain unfounded
% = do not bring disrepute upon AES
. 
The following information is based particularly on the original papers and the
articles \cite{cm:Wobst-iX2002} and \cite{cm:Lucks-DuD2002}.

@@

In their 2001-paper Ferguson, Schroeppel and Whiting \cite{cm:Ferguson2001}
presented a new method of symmetric codes cryptanalysis: They described AES with
a closed formula (in the form of a continued fraction) which was possible
because of the "relatively" clear structure of AES. This formula consists of
around 1000 trillion terms of a sum - so it does not help concrete practical
cryptanalysis. Nevertheless curiosity in the academic community was awakened.
It was already known, that the 128-bit AES could be described as an
over-determined system of about 8000 quadratic equations (over an algebraic
number field) with about 1600 variables (some of them are the bits of the wanted
key) -- equation systems of that size are in practice not solvable. This special
equation system is relatively sparse, so only very few of the quadratic terms
(there are about 1,280,000 are possible quadratic terms in total) appear in the
equation system.

The mathematicians Courtois and Pieprzyk \cite{cm:Courtois2002} published a paper
in 2002, which got a great deal of attention amongst the cryptology community: The
pair had further developed the XL-method (eXtended Linearization), introduced at
Eurocrypt 2000 by Shamir et al., to create the so called XSL-method (eXtended
Sparse Linearization). The XL-method is a heuristic technique, which in some
cases manages to solve big non-linear equation systems and which was till then
used to analyze an asymmetric algorithm (HFE).  The innovation of Courtois and
Pieprzyk was, to apply the XL-method on symmetric codes: the XSL-method can be
applied to very specific equation systems. A 256-bit AES could be attacked in
roughly $2^{230}$ steps. This is still a purely academic attack, but also a
direction pointer for a complete class of block ciphers. The major problem with
this attack is that until now nobody has worked out, under what conditions it is
successful: the authors specify in their paper necessary conditions, but it is
not known, which conditions are sufficient.  There are two very new aspects of
this attack: firstly this attack is not based on statistics but on algebra. So
attacks seem to be possible, where only very small amounts of ciphertext are
available. Secondly the security of a product algorithm%
\index{product algorithm}\index{encryption!product algorithm}%
\index{cascade cipher}\index{encryption!cascade cipher}%
\footnote{%
A ciphertext can be used as input for another encryption algorithm. 
A cascade cipher%
\index{product algorithm}\index{encryption!product algorithm}%
\index{cascade cipher}\index{encryption!cascade cipher}%
is build up as a composition of different encryption transformations.
The overall cipher is called product algorithm or cascade cipher
(sometimes depending whether the used keys are statistically dependent or not).\\
Cascading does not always improve the security.\\
This process is also used {\em within} modern algorithms:
They usually combine simple and, considered at its own, cryptologically
relatively insecure single steps in several rounds into an efficient 
overall procedure.  Most block ciphers (e.g. DES, IDEA) are cascade ciphers.\\
Also serial usage of the same cipher with different keys (like with Triple-DES)
is called cascade cipher. 
}
does not exponentially increase with the number of rounds.

Currently there is a large amount of research in this area: for example Murphy and Robshaw presented a paper at Crypto 2002 \cite{cm:Robshaw2002a}, which could dramatically improve cryptanalysis: the burden for a 128-bit key was estimated at about $2^{100}$ steps by describing AES as a special case of an algorithm called BES (Big Encryption System), which has an especially "round" structure. But even $2^{100}$ steps are beyond what is achievable in the foreseeable future. Using a 256 bit key the authors estimate that a XSL-attack will require $2^{200}$ operations.

@@

So for AES-256 the attack is much more effective than brute-force\index{attack!brute-force} but still far more away from any computing power which could be accessible in the short-to-long term.

The discussion is very controversial at the moment: Don Coppersmith (one of the
inventors of DES) for example queries the practicability of the attack because
XLS would provide no solution for AES \cite{cm:Coppersmith2002}. This implies that
then the optimization of Murphy and Robshaw \cite{cm:Robshaw2002b} would not work.

In 2009 Biryukov and Khovratovich \cite{cm:Biryukov2009} published another
theoretical attack on AES. This attack uses different methods from the ones
described above. They applied methods from hash function cryptanalysis (local
collisions and boomerang switching) to construct a related-key attack on
AES-256. I.~e.\ the attacker not only needs to be able to encrypt arbitrary data
(chosen plain text), in addition he needs to be able to manipulate the unknown key
(related-key).

@@

% --------------------------------------------------------------------------
\subsection{Algebraic or algorithmic cryptanalysis on symmetric algorithms}
\index{attack!algebraic}
\index{SAT solver}
\label{Algebraic-gegen-Symmetr}

@@

Example: The Boolean formula "A AND B" evaluates to 1, if and only if A=B=1. For the formula "A AND NOT(A)" there exists no assignment of its variable A, so that the formula is evaluated to the value 1.

For larger Boolean formulas, it is not easy to determine if an assignment exists for which the formula can be evaluated to 1 (this problem belongs to the NP-complete problems). Therefore specific tools have been developed to solve this problem for general Boolean formulas, so called SAT solvers\footnote{
    With {\bf CT2}\index{CT2}\index{CrypTool 2} you can execute
    a SAT solver -- using in the Startcenter
    {\bf Templates
    \textbackslash{} Mathematics \textbackslash{}
    SAT Solver (Text Input)}  and  {\bf SAT Solver (File Input)}.
    }. As has been found, SAT solvers can also be used to attack cryptographic systems.

@@

The general approach to use SAT solvers in cryptanalysis is very straightforward: First, the cryptographic problem, e.g. finding the symmetric key or an inversion of a hash function, is translated into a SAT problem. This is done in a way that the solution of the SAT problem solves the original cryptographic problem. Therefore, the SAT solver can be used to find a solution to the SAT problem. The paper by Massacci \cite{cm: Massacci2000} describes the first known usage of a SAT solver in this context. Unfortunately, very soon it turned out that such a general approach cannot be used efficiently in practice. This is due to the fact that the cryptographic SAT problems are very complex and the runtime of a SAT solver increases exponentially with the problem size. Therefore, in modern approaches SAT solvers are used only for solving partial problems of cryptanalysis. A good example for this is described in the paper by Mironov and Zhang \cite{cm: Mironov2006}. They demonstrate the usage of a SAT solver in an attack on hash functions, where the SAT solver is used to solve some partial problems in a very efficient way.

xxxxxxxxx Bsp aus CT2.1 ergänzen
%\index{ZZZ6b-xxxxxxxxxxxxxxxx_Bsp aus CT2.1 ergaenzen}
\index{ZZZ6b-xxxxxxxxxxxxxxxx}

% --------------------------------------------------------------------------
\subsection{Current status of brute-force attacks on symmetric algorithms}
\index{attack!brute-force}
\index{RC5}
\label{Brute-force-gegen-Symmetr}

@@

\noindent Below is SageMath source code related to contents of the
chapter~\ref{Label_Kapitel_1} (``\nameref{Label_Kapitel_1}'').

Further details concerning cryptosystems within SageMath (e.g. S-DES) can be
found e.g. in the thesis of Minh Van Nguyen \cite{cm:Nguyen2009}.

@@

Mini-AES, originally described at \cite{cm:Phan2002}, is a simplified variant of the
Advanced Encryption Standard (AES) to be used for cryptography education.

>>>>

% $Id$
% ..............................................................................
%                      E n c r y p t i o n   M e t h o d s
% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

%% Remark: Mit \begin{bibunit}[unsrt]  hier würde ein Kapitel-Bibliography erzeugt, das
%%         Zahlen anzeigt und unsortiert ist. Die Reihenfolge ergibt dann sich anhand der
%%         Abfolge der Referenzen im Quelltext.
\begin{bibunit}[babalpha]  %% alpha: Kapitel-Bibliography zeigt Autorenkürzel und ist auch danach
	                   %%     sortiert. Gleiches Kürzel wie in der zweiten Gesamtbibliography.

\hypertarget{Chapter_EncryptionSecDefinitions}{}

\chapter{Security Definitions and Encryption Procedures}
\label{Chapter_EncryptionSecDefinitions}
(\hyperlink{author_Bernhard-Esslinger}{Bernhard Esslinger},
 \hyperlink{author_Joerg-Cornelius-Schneider}{Joerg-Cornelius Schneider},
 May 1999; Updates Dec 2001, Feb 2003, Jun 2005, Jul 2007, Jan 2010, Mar 2013, Aug 2016)

@@

The purpose of encryption \index{encryption} is to change data in such a way
that only an authorized recipient is able to reconstruct the plaintext. This
allows us to transmit data without worrying about it getting into unauthorized
hands. Authorized recipients possess a piece of secret information -- called the
key -- which allows them to decrypt the data while it remains hidden from
everyone else.\footnote{%
  However, an attacker still can disturbe the connection or tap metadata (like who
  is communicating with whom).
}

For explanations in the following we use the notation from 
Figure \ref{cm_Generic-Notations-when-Encrypting}:
\begin{figure}[ht]
\begin{center}
\includegraphics[scale=0.7]{figures/Generic-Notation-Encryption_en.png}
\caption{Common notations when using ciphers} 
\label{cm_Generic-Notations-when-Encrypting}
\end{center}
\end{figure}

% --------------------------------------------------------------------------
\hypertarget{cm_Section_Security_Definitions}{}
% \section[Security definitions]{Security definitions}
\section{Security definitions and the importance of cryptology}
\label{cm_Section_Security_Definitions}
\index{security definitions}

@@

@@

\vskip +3pt
Depending on the adversary's capabilities there are mainly two basic notations
of security distinguished in literature
(see e.g. {\em Contemporary Cryptography} \cite{Oppliger2011}):

@@

\item {\bf Computational, conditional or practical security}\\
  A cipher is {\em computationally} secure if it is theoretically possible to break
  such a system but it is infeasible to do so by any known practical means. Theoretical
  advances (e.g., improvements in integer factorization algorithms) and faster
  computing technology require these solutions to be continually adapted.

@@

\item {\bf Information-theoretical or unconditional security}\\
  A cipher is considered {\em unconditionally} secure if its security is guaranteed
  no matter how much resources (time, space) the attacker has -- so even in the case
  where the adversary has unlimited resources for breaking a cipher. Even with
  unlimited resources an adversary is unable to gain any meaningful data from a
  ciphertext.

  There exist information-theoretically secure schemes that provably cannot be
  broken even with unlimited computing power -- an example is the
  {\em one-time pad} (OTP\index{OTP}).

@@

\item {\bf Ad-hoc security}
A cryptographic system has this security feature if it is not worth to try to
break into such a system because of inadequate price of data with comparison to
price of work needed to do so. Or an attack can't be done in sufficiently short
time (see {\em Handbook of Applied Cryptography} \cite{Menezes2001}).

@@

\vskip +3pt
For good procedures used today the time taken to break them is so long that
it is practically impossible to do so, and these procedures can therefore be
considered (practically) secure -- from a pure algorithm's point of view.\footnote{%
  Especially after the knowledge gathered by Edward Snowden\index{Snowden, Edward}
  there were many discussions,
  whether encryption is secure. In \cite{Esslinger2014} is the result of an evaluation,
  which cryptographic algoritms can be relied on -- according to current knowledge.
  The article investigates: Which crypto systems can -- despite the reveal of
  the NSA/GCHQ attacks -- still be considered as secure? Where have systems
  been intentionally weakened? How can we create a secure cryptographic future?
  Where is the difference between maths and implementation?
}

\vskip +3pt
We basically distinguish between symmetric
(see chapter \ref{cm_Section_Symmetric-encryption}) and asymmetric
(see chapter \ref{cm_Section_Asymmetric-encryption}) encryption
procedures.
The books of Bruce Schneier \cite{Schneier1996} and Klaus Schmeh
\cite{Schm2016} also offer a very good overview of the different
encryption algorithms.\footnote{%
  A compact overview about what is used where, which methods are secure, where you
  have to anticipate problems and where the construction areas will be in the
  future (incl. the lengthy procedures of the standardization) can be found in
  the German article \cite{Schmidt2016}.
}

\vskip +3pt
With the use of the internet and wireless communication, encryption technologies
are used (mostly transparently) by everyone. However, they have been in use since
centuries by governments, military, and diplomats. The side which had a better command
of these technologies could exert big influence on {\bf politics} and war with the help
of secret services.
This book only touches history when introducing the earlier cipher methods in
chapter \ref{Chapter_PaperandPencil}. You can gain an impression, how important
cryptology was and still is, by considering the following two examples: the
educational film ``War of the letters'' (German: Krieg der Buchstaben)\footnote{%
  See \url{http://bscw.schule.de/pub/bscw.cgi/d1269787/Krieg_der_Buchstaben.pdf}.\\
  A supporting quote from Denis Smyth, professor in the Department of History
  and the International Relations Programme at the University of Toronto:
  ``Secret intelligence has long been regarded as the ``missing dimension'' of
  international relations.'' (from \url{http://www.secretintelligencefiles.com})
}
and the debates around the so called crypto wars\footnote{%
  See \url{https://en.wikipedia.org/wiki/Crypto_Wars}.\index{crypto wars}
}.

@@

\newpage
The following figure \ref{cm_Figure_OTP-demo-pictures} intends to give an idea, that it is impossible to determine the correct plaintext from a OTP\index{OTP} (if the OTP method has been applied correctly and if all keys have the same likelyhood).

The example in this figure uses an 8 character long given ciphertext:
\texttt{11 1B 1E 18 00 04 0A 15}.
There are many meaningful words with 8 letters and for each there is an according key.
So an attacker can not determine alone from the ciphertext, which is the correct key and which is the correct plaintext word.

Also see figure~\ref{fig-bool-otp} in chapter~\ref{s-bool-bitstr-real-random},
where an according example with text strings is build with SageMath.
\begin{figure}[ht]
\begin{center}
\includegraphics[scale=0.55]{figures/OTP-demo-pictures-en.png}
\caption[Illustration for the information-theoretically secure OTP scheme]
        {Illustration for the information-theoretically secure OTP scheme\footnotemark}
\label{cm_Figure_OTP-demo-pictures}
\end{center}
\end{figure}
\footnotetext{
  Picture source: Free pictures from \url{https://pixabay.com/}
}

@@

% --------------------------------------------------------------------------
\section[Symmetric encryption]
        {Symmetric encryption\footnotemark}
  \footnotetext{%
    With CrypTool~1 ({\bf CT1})\index{CT1} you can execute
    the following modern symmetric encryption algorithms 
    (using the menu path {\bf Crypt \textbackslash{} Symmetric (modern)}):\\
    IDEA, RC2, RC4, DES (ECB), DES (CBC), Triple-DES (ECB), Triple-DES (CBC),
    MARS (AES candidate), RC6 (AES candidate), Serpent (AES candidate), 
    Twofish (AES candidate), Rijndael (official AES algorithm)\index{AES}.\\
    With CrypTool~2 ({\bf CT2})\index{CT2} you can execute the
    following modern symmetric encryption algorithms (using in the Startcenter
    {\bf Templates
    \textbackslash{} Cryptography \textbackslash{} Modern \textbackslash{}
    Symmetric}):\\
    AES, DES, PRESENT, RC2, RC4, SDES, TEA, Triple-DES, Twofish.\\
    In JCrypTool ({\bf JCT})\index{JCT} you can execute the
    following modern symmetric encryption algorithms:\\ 
    AES, Rijndael, Camellia, DES, Dragon, IDEA, LFSR, MARS, Misty1, RC2, RC5,
    RC6, SAFER+, SAFER++, Serpent, Shacal, Shacal2, Twofish.
  }
\label{cm_Section_Symmetric-encryption}

@@

This is visualized in Figure \ref{cm_Figure_Symmetric-Enc_Secret-Key-Enc}:
\begin{figure}[ht]
\begin{center}
\includegraphics[scale=0.7]{figures/SymmetricEnc_Figure_Chap1_en.png}
\caption{Symmetric or secret-key encryption} 
\label{cm_Figure_Symmetric-Enc_Secret-Key-Enc}
\end{center}
\end{figure}

All classical ciphers are of the type symmetric. Examples can be found within
the CT programs, in chapter~\ref{Chapter_PaperandPencil}
(``\nameref{Chapter_PaperandPencil}'')
of this script, or in \cite{Nichols1996}.
In this section however, we want to consider only modern symmetric mechanisms.

@@

\subsection[AES (Advanced Encryption Standard)]{AES (Advanced Encryption Standard)\footnotemark} 
\footnotetext{%
   In CT1\index{CT1} you can find 3 visualizations for this cipher via
   the menu {\bf Indiv. Procedures \textbackslash{} Visualization of
   Algorithms \textbackslash{} AES}.\\
   In CT2\index{CT2} you can find a template performing AES step-by-step
   (by entering the search string ``AES'' in the Startcenter).
}
\label{CM_AES}
\index{AES}

Before AES, the most well-known modern symmetric encryption procedure was the
DES\index{DES} algorithm.
The DES algorithm has been developed by IBM in collaboration with the
National Security Agency \index{NSA} (NSA), and was published as a standard in
1975. Despite the fact that the procedure is relatively old, no effective attack
on it has yet been detected. The most effective way of attacking consists of
testing (almost) all possible keys until the right one is found ({\em brute-force-attack}).
\index{attack!brute-force} Due to the relatively short key length of
effectively 56 bits (64 bits, which however include 8 parity bits), numerous
messages encrypted using DES have in the past been broken. Therefore, the
procedure can not be considered secure any longer. Alternatives to the DES
procedure include IDEA\index{IDEA}, Triple-DES (TDES) and especially AES.
\par \vskip + 3pt

@@

An introduction and further references about the AES algorithms and the AES
candidates of the last round can be found i.e. within the online help of
CrypTool\index{CrypTool}%
\footnote{%
      CrypTool~1 online help\index{CT1}: The index head-word {\bf AES}
      leads to the 3 help pages: {\bf AES candidates}, 
      {\bf The AES winner Rijndael} and 
      {\bf The Rijndael encryption algorithm}\\
      A comprehensive description of AES including C code can be found
      in \cite{Haan2008}.
  }
oder in Wikipedia%
\footnote{%
	\url{https://en.wikipedia.org/wiki/Advanced_Encryption_Standard}
  }.

\clearpage
\noindent The two screenshots \ref{AES-Visualization-Zabala-Flash-step3} and
\ref{AES-Visualization-Zabala-Flash-step5} are taken from one of three
AES visualizations in CT1\index{CT1}.

\begin{figure}[!ht]
\begin{center}
\includegraphics[scale=0.6]{figures/AES-Visualization-Zabala-Flash-step3_en}
\caption{AES visualization by Enrique Zabala from CT1 (part 1)}
\label{AES-Visualization-Zabala-Flash-step3}
\end{center}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics[scale=0.6]{figures/AES-Visualization-Zabala-Flash-step5_en}
\caption{AES visualization by Enrique Zabala from CT1 (part 2)}
\label{AES-Visualization-Zabala-Flash-step5}
\end{center}
\end{figure}
%\end{sloppypar}

\clearpage
\noindent Now we want to encrypt with AES in CBC mode a 128-bit block of plaintext.
Off the resulting ciphertext we are only interested in
the 1st block (if there is more, it would be padding, here null-padding).
For demonstration we do it once with CT2 and once with OpenSSL.\\

\noindent Figure \ref{AES_Encrypting-one-block-with-CT2} shows the encryption of
one block in {\bf CT2}\index{CT2}.

\noindent The plaintext ``AESTEST1USINGCT2'' is converted to hex
(41 45 53 54 45 53 54 31 55 53 49 4E 47 43 54 32). Using this and the key
3243F6A8885A308D313198A2E0370734 the AES component creates the ciphertext.
It is in hex:\\
B1 13 D6 47 DB 75 C6 D8 47 FD 8B 92 9A 29 DE 08

\begin{figure}[ht]
\begin{center}
\includegraphics[scale=0.42]{figures/AES_Encrypting-one-block-with-CT2_en}
\caption{AES encryption (of exactly 1 block and without padding) in CT2} 
\label{AES_Encrypting-one-block-with-CT2}
\end{center}
\end{figure}

\vspace{15pt}
\noindent The same result can be achieved with {\bf OpenSSL}\index{OpenSSL}%
\footnote{%
      OpenSSL\index{OpenSSL} is a very widespread free open-source crypto library,
      used by many applications, for instance to implement the TLS protocol.
      Part of OpenSSL is the commandline tool openssl, which can be used to test
      the functionality directly on many operating systems and to request, create
      and manage certificates.\\
      Contrarily to the also very widespread and very good commandline tool gpg from
      GNU Privacy Guard\index{GnuPG}
      (\url{https://en.wikipedia.org/wiki/GNU_Privacy_Guard}),
      openssl also allows calls with many details.
      The gpg has its focus on the practically applied ciphersuites.
      As far as we know, it is not possible to encrypt just one block without padding
      with the commandline tool gpg.\\
      Also see \url{https://en.wikipedia.org/wiki/OpenSSL}.
  }
from the commandline:
\begin{opensslcode}
\begin{Verbatim}%
[fontsize=\footnotesize]
>openssl enc -e -aes-128-cbc
        -K 3243F6A8885A308D313198A2E0370734
        -iv 00000000000000000000000000000000
        -in klartext-1.hex  -out klartext-1.hex.enc
>dir
06.07.2016  12:43                16 key.hex
20.07.2016  20:19                16 klartext-1.hex
20.07.2016  20:37                32 klartext-1.hex.enc
\end{Verbatim}
\caption{AES encryption (of exactly one block and without padding) in OpenSSL}
\label{cm_AES_no-padding:OpenSSL_example}
\end{opensslcode}

\clearpage
% HACK to fix warning "destination with the same identifier .. has already been used, ...":
%\makeatletter \renewcommand{\thepage}{\csname @arabic\endcsname \c@page} \makeatother
% --------------------------------------------------------------------------
\subsubsection{Results about theoretical cryptanalysis of AES}
\label{cm_New-AES-Analysis}
\index{cryptanalysis}
\index{AES}

Below you will find some results, which have recently called into question the security of the AES algorithm -- from our point of view these doubts practically still remain unfounded
% = do not bring disrepute upon AES
. 
The following information is based particularly on the original papers and the
articles \cite{Wobst2002} and \cite{Lucks2002}.

@@

In their 2001-paper Ferguson, Schroeppel and Whiting \cite{Ferguson2001}
presented a new method of symmetric codes cryptanalysis: They described AES with
a closed formula (in the form of a continued fraction) which was possible
because of the ``relatively'' clear structure of AES. This formula consists of
around 1000 trillion terms of a sum - so it does not help concrete practical
cryptanalysis. Nevertheless curiosity in the academic community was awakened.
It was already known, that the 128-bit AES could be described as an
over-determined system of about 8000 quadratic equations (over an algebraic
number field) with about 1600 variables (some of them are the bits of the wanted
key) -- equation systems of that size are in practice not solvable. This special
equation system is relatively sparse, so only very few of the quadratic terms
(there are about 1,280,000 are possible quadratic terms in total) appear in the
equation system.

The mathematicians Courtois and Pieprzyk \cite{Courtois2002} published a paper
in 2002, which got a great deal of attention amongst the cryptology community: The
pair had further developed the XL-method (eXtended Linearization), introduced at
Eurocrypt 2000 by Shamir et al., to create the so called XSL-method (eXtended
Sparse Linearization). The XL-method is a heuristic technique, which in some
cases manages to solve big non-linear equation systems and which was till then
used to analyze an asymmetric algorithm (HFE).  The innovation of Courtois and
Pieprzyk was, to apply the XL-method on symmetric codes: the XSL-method can be
applied to very specific equation systems. A 256-bit AES could be attacked in
roughly $2^{230}$ steps. This is still a purely academic attack, but also a
direction pointer for a complete class of block ciphers. The major problem with
this attack is that until now nobody has worked out, under what conditions it is
successful: the authors specify in their paper necessary conditions, but it is
not known, which conditions are sufficient.  There are two very new aspects of
this attack: firstly this attack is not based on statistics but on algebra. So
attacks seem to be possible, where only very small amounts of ciphertext are
available. Secondly the security of a product algorithm%
\index{product algorithm}\index{encryption!product algorithm}%
\index{cascade cipher}\index{encryption!cascade cipher}%
\footnote{%
A ciphertext can be used as input for another encryption algorithm. 
A cascade cipher%
\index{product algorithm}\index{encryption!product algorithm}%
\index{cascade cipher}\index{encryption!cascade cipher}%
is build up as a composition of different encryption transformations.
The overall cipher is called product algorithm or cascade cipher
(sometimes depending whether the used keys are statistically dependent or not).\\
Cascading does not always improve the security.\\
This process is also used {\em within} modern algorithms:
They usually combine simple and, considered at its own, cryptologically
relatively insecure single steps in several rounds into an efficient 
overall procedure.  Most block ciphers (e.g. DES, IDEA) are cascade ciphers.\\
Also serial usage of the same cipher with different keys (like with Triple-DES)
is called cascade cipher.\index{DES!Triple-DES}
}
does not exponentially increase with the number of rounds.

Currently there is a large amount of research in this area: for example Murphy and Robshaw presented a paper at Crypto 2002 \cite{Robshaw2002a}, which could dramatically improve cryptanalysis: the burden for a 128-bit key was estimated at about $2^{100}$ steps by describing AES as a special case of an algorithm called BES (Big Encryption System), which has an especially ``round'' structure. But even $2^{100}$ steps are beyond what is achievable in the foreseeable future. Using a 256 bit key the authors estimate that a XSL-attack will require $2^{200}$ operations.

@@

So for AES-256 the attack is much more effective than brute-force\index{attack!brute-force}
but still far away from any computing power which could be accessible in the short-to-long term.

The discussion is very controversial at the moment: Don Coppersmith (one of the
inventors of DES) for example queries the practicability of the attack because
XLS would provide no solution for AES \cite{Coppersmith2002}. This implies that
then the optimization of Murphy and Robshaw \cite{Robshaw2002b} would not work.

In 2009 Biryukov and Khovratovich \cite{Biryukov2009} published another
theoretical attack on AES. This attack uses different methods from the ones
described above. They applied methods from hash function cryptanalysis (local
collisions and boomerang switching) to construct a related-key attack on
AES-256. I.~e.\ the attacker not only needs to be able to encrypt arbitrary data
(chosen plain text), in addition he needs to be able to manipulate the unknown key
(related-key).

@@

% --------------------------------------------------------------------------
\subsection{Algebraic or algorithmic cryptanalysis on symmetric algorithms}
\index{attack!algebraic}
\index{SAT solver}
\label{cm_Algebraic-versus-Symmetr}

@@

Example: The Boolean formula ``A AND B'' evaluates to 1, if and only if A=B=1. For the formula ``A AND NOT(A)'' there exists no assignment of its variable A, so that the formula is evaluated to the value 1.

For larger Boolean formulas, it is not easy to determine if an assignment exists for which the formula can be evaluated to 1 (this problem belongs to the NP-complete problems). Therefore specific tools have been developed to solve this problem for general Boolean formulas, so called SAT solvers\footnote{
    With {\bf CT2}\index{CT2} you can execute
    a SAT solver -- using in the Startcenter
    {\bf Templates
    \textbackslash{} Mathematics \textbackslash{}
    SAT Solver (Text Input)}  and  {\bf SAT Solver (File Input)}.
    }. As has been found, SAT solvers can also be used to attack cryptographic systems.

@@

The general approach to use SAT solvers in cryptanalysis is very straightforward: First, the cryptographic problem, e.g. finding the symmetric key or an inversion of a hash function, is translated into a SAT problem. Then, the SAT solver can be used to find a solution to the SAT problem. The solution of the SAT problem then also solves the original cryptographic problem.
The paper by Massacci \cite{Massacci2000} describes the first known usage of a SAT solver in this context. Unfortunately, very soon it turned out that such a general approach cannot be used efficiently in practice. This is due to the fact that the cryptographic SAT problems are very complex and the runtime of a SAT solver increases exponentially with the problem size. Therefore, in modern approaches SAT solvers are used only for solving partial problems of cryptanalysis. A good example for this is described in the paper by Mironov and Zhang \cite{Mironov2006}. They demonstrate the usage of a SAT solver in an attack on hash functions, where the SAT solver is used to solve some partial problems in a very efficient way.

% --------------------------------------------------------------------------
\subsection{Current status of brute-force attacks on symmetric algorithms}
\index{attack!brute-force}
\index{RC5}
\label{cm_Brute-force-versus-Symmetr}

@@

\noindent Below is SageMath source code related to contents of the
chapter~\ref{CM_Analysis-SymCiphers-Educational}
(``\nameref{CM_Analysis-SymCiphers-Educational}'').

Further details concerning cryptosystems within SageMath (e.g. about the Simplified
Data Encryption Standard SDES) can be found e.g. in the thesis of Minh Van
Nguyen \cite{Nguyen2009b}.

@@

Mini-AES, originally described at \cite{Phan2002}, is a simplified variant of the
Advanced Encryption Standard (AES) to be used for cryptography education.
